% ------------------------------------------------------------------------
% -*-TeX-*- -*-Hard-*- Smart Wrapping
% ------------------------------------------------------------------------
% Thesis Acknowledgements ------------------------------------------------

\addtolength{\topmargin}{-.875in}

\prefacesection{Acknowledgements}
%\def\baselinestretch{1.0}
\setlinespacing{1.66}

\medskip
\par
I would like to thank my supervisors Prof Karim Djouani and Prof Yskandar
Hamam for their guidance and support. From the very beginning, their
experience and wisdom have been instrumental for the completion of this work.

\medskip

My gratitude to Tshwane University of Technology (TUT) and the French South
African Institute of Technology (F'SATI) for providing a suitable research
environment. My gratitude is particularly extended to Dr Anish Kurien, for
his support and constructive advice.

\medskip

I am profoundly thankful to the \emph{Laboratoire d'Ing\'{e}nierie des
Syst\`{e}mes de Versailles} (LISV) for providing the necessary research
material and environment for the experiments conducted. A special word of
thanks to Prof Eric Monacelli, Prof Sylvain Chevallier, and Dr Zoran Tiganj.
They have made my stay in the laboratory a success, and this work could not
have been possible without their assistance.

\medskip

I express my sincere gratitude to the Mandela-Rhodes Foundation (MRF) for its
financial support throughout my entire masters studies. When I had to join
the LISV, the Foundation did not hesitate to support the ticket fare from
South Africa to France and back.

% ----------------------------------------------------------------------
